---
title: "Estimating transmission impact on vehicle fuel economy"
author: "Syed Abdullah Hasan"
date: "7/15/2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
cache       = TRUE, fig.width   = 8, fig.height  = 4, fig.align   = 'center',
echo        = F, message     = F, include     = T, warning     = F)

```

```{r Loading_Data_and_Libraries, include = F}
library (dplyr)
library (reshape2)
library (MASS)
library (ggplot2)
library (scales)
library (kableExtra)
library (olsrr)
```

# Executive Summary  
This project studies the impact of vehicle transmission on fuel economy by examining a sample of 32 vehicles from 1973-74 model years studied by Motor Trend US Magazine. Initial exploratory analysis of the dataset examines impact of transmission on fuel economy as well as underlying correlations between all variables in the dataset. Subsequently an iterative backward selection approach is undertaken to select the optimum regression model, in which vehicle quarter mile times and weight are identified as the key predictors alongside the transmission. Finally, the report concludes that manual transmission vehicles are generally better for fuel economy as compared with automatic transmission vehicles, and add 2.94 miles per gallon on average to fuel economy after adjusting for vehicle performance (proxied by vehicle weight and quarter mile times).

# Methodology
This report analyses data collected by Motor Trend US Magazine in 1974 to evaluate the impact of vehicle transmission fuel economy, by considering 10 aspects of automobile design & performance for a set of 32 automobiles (produced in the year 1973-74). The study will first explore key features of the data set, proceed to fit an appropriate regression model and subsequently evaluate the following questions:

1. Is an automatic or manual transmission better for fuel economy measured in MPG (miles per gallon)?

2. What is the difference in MPG between automatic and manual transmission vehicles?

# Exploratory Data Analysis
```{r, include=F}
data(mtcars)
mtcars$am<-as.factor(mtcars$am)
mtcars$vs<-as.factor(mtcars$vs)
```
The data set used for the report includes 32 observations on the following set of numeric variables:  
  1. `mpg`	      - Miles/(US) gallon  
  2. `cyl`	      - Number of cylinders  
  3. `disp`	      - Displacement (cu.in.)  
  4. `hp`	        - Gross horsepower  
  5. `drat`	      - Rear axle ratio  
  6. `wt`	        - Weight (1000 lbs)  
  7. `qsec`	      - 1/4 mile time  
  8. `vs`	        - Engine (0 = V-shaped, 1 = straight)  
  9. `am`	        - Transmission (0 = automatic, 1 = manual)  
  10. `gear`      - Number of forward gears  
  11. `carb`      - Number of carburetors  

 * Note: Variables `vs` and `am` are converted to factors prior to model specification.

As seen in Chart 1 (Appendix), `mpg` is clearly correlated to each of the variables in the data set, hence all variables should be evaluated in model selection when analyzing the isolated impact of transmission on fuel economy. 

Chart 2 shows that automatic transmission vehicles have a better fuel economy on average as compared with manual transmission vehicles. By specifying a base regression model for `mpg` against `am` (Table 1), we can confirm that the average fuel economy (`mpg`) for manual transmission vehicles is `r round(mean(subset(mtcars, am==1)[,1]),2)` whereas that for automatic transmission vehicles is `r round(mean(subset(mtcars, am==0)[,1]),2)`. Both coefficients in the model are statistically significant at an alpha level of 5%.

However, this analysis does not account for the impact of additional variables in the data set on fuel economy. From Table 2, we can observe a high correlation between each of the  variables in the data set with `mpg` and with each other. Correlated predictors may introduce multicollinearity in the model, confounding interpretation of potentially significant predictors. Hence, we need to be careful not to over-fit variables in the best-fit model.

# Model Specification and Selection

Step-wise regression is employed to remove predictors in each iteration from a fully specified model until the best performing model with the lowest prediction error is determined. The Akaike Information Criterion (AIC) is used to evaluate the model with the best fit. 

The full model includes `am` as a factor variable with the intercept term suppressed. This model structure allows to capture impact of transmission (`am`) on fuel economy (`mpg`) via the intercept term. Table 3 shows that none of the predictors are statistically significant in this model.

``` {r}

step.model <- stepAIC (full.model, direction = "both", trace = F)
```

The model with the lowest AIC which provides the best fit for the data is as follows:
$$mpg =  \beta_0 + \beta_1 * wt + \beta_2 * qsec$$
where $$\beta_0$$ is the intercept capturing impact of transmission `am` on fuel economy `mpg`.

## Analysis of Residuals

The selected model residuals are now screened to rule out any evidence of heteroskedasticity and influential outliers. The following observations confirm that the variance of residuals is constant:  

 * The plot of residuals against fitted-vales (Chart 3) for the best-fit model shows a largely random distribution. 
 * The Normal Q-Q plot shows majority of points falling along the dotted line, indicating that the residuals are normally distributed.
 * The scale-location plot confirms a normal distribution. 
 * Despite the presence of outliers (Chart 4), the Cook's D-bar plot shows that all distances are significantly less than 1.
 * We can formally test for heteroskedasticity using the Breush Pagan test and the F-test (Table 5). Both tests show that the null hypothesis cannot be rejected, we proceed to select the current model and confirm that heteroskedasticity is not significantly affecting the interpretation of the model variables.

# Model Interpretation  
The following conclusions may be drawn from the best-fit model selected in this analysis:  

 * The model explains `r round(summary(step.model)$adj.r.squared *100,2)` percent of variation in the outcome variable `mpg`.
 * The model predicts that an automatic transmission `am = 1` increases fuel economy (`mpg`) on average by  `r round(summary(step.model)$coef[4,1] -  summary(step.model)$coef[3,1],2)` miles per gallon, after adjusting for the impact of weight and quarter mile times of sampled cars.
 * Based on the model findings, automatic transmission vehicles are generally better for fuel economy as compared with manual transmission vehicles.

\newpage

# Appendix
## Dataset Summary
```{r, include = T}
summary (mtcars) 
```

## Charts
``` {r Charts_and_Tables}
# Scatterplot showing relationship between mpg and all other variables
data (mtcars)
mtcars3 <- cbind("cars"=row.names(mtcars),mtcars)
mtcars3 <- melt(mtcars3, id.vars = c("cars","mpg"))
g1 <- ggplot (mtcars3, aes(x=mpg)) + 
        facet_wrap (~variable, scales = "free_y") +
        geom_point (aes(y=value, col=variable)) + 
        theme (legend.position = "none") +
        scale_x_continuous (n.breaks = 4) + 
        scale_y_continuous (guide = guide_axis(check.overlap = TRUE))

# Boxplot of mpg against am
g2 <- ggplot (mtcars, aes(x=as.factor(am), y=mpg)) +
        geom_boxplot (aes(fill=am)) +
        geom_jitter (position = position_jitter(0.2),
                     alpha = 0.8, fill = "white") +
        labs (x = "Automatic Transmission") + 
        theme (legend.position = "none")

```
### Chart 1 - Scatter plot of MPG against all other variables  
```{r, include=T}
g1
```

### Chart 2 - Boxplot of MPG against AM  
```{r, include =T}
g2
```

### Chart 3 - Residual plots for Best-fit Model 
```{r, include=T}
par(mfrow=c(2,2), mar=c(2,2,2,2), cex=0.5)
plot(step.model)
```

### Chart 4 - Cooks D-bar plot for Best-fit Model 
```{r, include=T}
par(mfrow=c(1,1), mar=c(2,2,2,2), cex=0.5)
ols_plot_cooksd_bar(step.model)
```

## Tables  
### Table 1 - Summary Table - Full Regression Model 
``` {r, include =T}
base.model <- lm(mpg~factor(am),mtcars)
summary (base.model)
```
### Table 2 - Correlation matrix for data set  
```{r, include=T}
data (mtcars)
mtcor <- round (cor (mtcars),2)
upper <- mtcor
upper [upper.tri(mtcor)] <- ""
upper <- as.data.frame (upper)
upper
```
### Table 3 - Summary Table - Full Regression Model 
``` {r, include =T}
data(mtcars)
mtcars$am <- as.factor(mtcars$am)
mtcars$vs<-as.factor(mtcars$vs)
full.model <- lm (mpg~. -1,data=mtcars)
summary (full.model)
```
### Table 4 - Summary Table - Best-fit Regression Model 
``` {r, include =T}
summary (step.model)
```
### Table 5 - Breusch-pagan test for Heteroskedasticity
``` {r, include =T}
ols_test_breusch_pagan (step.model, rhs = TRUE)
ols_test_f (step.model)
```